Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-10 draft.tex
.

Equation \eqref{simple_NO_dyn_with_diff_equil} can be rewritten as
\begin{equation}
\left(\nabla^2 + \left( i\sqrt{\frac{\lambda}{D}}\right)^2\right) NO = \sum_{i} \delta^2(\mathbf{x}-\mathbf{x}_{neur,i})\cdot \frac{- \gamma \cdot r_i}{D}
\label{simple_NO_dyn_with_diff_equil_helmholtz}
\end{equation}
which is a two-dimensional Helmholtz equation with a superposition of (scaled) Dirac-functions. Thus, the solution of $\mathrm{NO}$ is composed of a superposition of shifted and scaled versions of the Green's function of the differential operator on the left of the equation. For each delta function $\delta^2(\mathbf{x}-\mathbf{x}_i)$, the solution is
\begin{equation}
NO_i(\mathbf{x}) = \frac{r_i \gamma}{2 \pi D} K_0 \left(|\mathbf{x}-\mathbf{x}_{neur,i}|\sqrt{\frac{\lambda}{D}} \right) \equiv r_i \cdot \psi_{point}(|\mathbf{x}-\mathbf{x}_{neur,i}|)
\label{solution_diff_equil_bessel}
\end{equation}
where $\mathrm{K_0}$ is the zeroth modified Bessel function of the second kind. This solution reveals a fundamental problem of modelling the sources of $\mathrm{NO}$-production as point sources: the fact that $\mathrm{K_0(x)}$ diverges to infinity for $\mathrm{x\rightarrow 0}$. It is merely due to the finite density of the numeric grid used for the simulation of the diffusion that allows for a finite target value of concentration. Note that this problem only occurs in the two- or three-dimensional version of the differential equation, whereas in one dimension, the fundamental solution can be expressed as an exponential function with respect to the distance to the origin, resulting in a well-defined finite value at $\mathrm{x=0}$.

Generally speaking, no matter how the actual shape of the numeric solution in the equilibrium at a constant production rate looks like, it is expected to be of the form
\begin{align}
NO_i(\mathbf{x}) &= r_i \cdot \psi (d(\mathbf{x}_{neur,i},\mathbf{x})) \label{general_diff_interaction} \\
d(\mathbf{x},\mathbf{y}) &\equiv |\mathbf{x}-\mathbf{y}| \; . \label{eucl_dist}
\end{align}
The full solution is then
\begin{equation}
NO(\mathbf{x}) = \sum_i NO_i(\mathbf{x})\;.
\label{full_sol_diff_equil}
\end{equation}
If we define
\begin{equation}
\psi_{ij} \equiv \psi_{ij} \equiv \psi (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i}))
\label{interact_matrix_elements}
\end{equation}
we can express the condition \eqref{NO_equil_cond} as
\begin{equation}
\sum_j \psi_{ij}\cdot r_j = NO_0
\label{NO_equil_cond_interact_matrix}
\end{equation}
or, as an operator
\begin{align}
\hat{\psi}\mathbf{r} &= NO_0 \mathbf{n} \label{NO_equil_cond_interact_matrix_operator} \\
\mathbf{n}&\equiv (1,1,...,1) \; .
\end{align}
The problem of finding the steady-state solution of the homeostatic constraint thus reduces to inverting $\hat{\psi}$:
\begin{equation}
\mathbf{r} = NO_0 \hat{\psi}^{-1} \mathbf{n}
\label{NO_euqil_cond_interact_matrix_operator_solve}
\end{equation}
Still, to acquire any prediction from this model, we had to find a modified, non-diverging version of $\mathrm{\psi (d(\mathbf{x}_{neur,i},\mathbf{x}))}$ that retains the shape given by \eqref{solution_diff_equil_bessel} for larger distances but approaches the correct "numeric error"-value at the origin, determined by the spacing of the numeric grid. We "solved" this problem by the following expression:
\begin{equation}
\psi_{approx.} \equiv \frac{1}{\left(\frac{1}{\psi_0^\varepsilon} + \frac{1}{\psi_{point}^\varepsilon}\right)^{\frac{1}{\varepsilon}}}
\label{Numeric_Solution_Expression_Trick}
\end{equation}
where $\mathrm{\varepsilon}$ determines the "smoothness" of transition between $\mathrm{\psi}$ and the cutoff value $\mathrm{\psi_0}$. We chose $\varepsilon=10$ for all further calculations. To find an expression for $\mathrm{\psi_0}$, we took the simple approach to interpret this value as a mean of the analytic solution across the square area covered by the corresponding grid cell. As an additional simplification, we substitute the necessary integration over the square grid cell by a circular area of equal size around the source. This calculation yields
\begin{equation}
\psi_0 = \frac{1-h\sqrt{\frac{\lambda}{\pi D}} K_1\left( h\sqrt{\frac{\lambda}{\pi D}}\right) }{h^2 \lambda}
\label{Numeric_Grid_Bessel_Approx}
\end{equation}
where $\mathrm{h}$ is the spatial resolution of the grid cells. Figure \ref{Psi_Approximation_Compare} shows a comparison between the numerically calculated solution and the expression given by \ref{Numeric_Solution_Expression_Trick}. As expected, the approximation fits very well for nonzero values. The value at the critical point at the origin is slightly underestimated by \eqref{Numeric_Grid_Bessel_Approx}, but nonetheless fits well into the overall shape.
\begin{figure}
%% compare_num_diff_steady_vs_analytic.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/Psi_Interact_Approximation.png}
\end{center}
\caption{Numerical steady-state solution of Equation \eqref{simple_NO_dyn_with_diff_equil_helmholtz} (blue, $\mathrm{h=10\;\mu m}$) and its approximation (dashed) given by \eqref{Numeric_Solution_Expression_Trick}. The numerical simulation was carried out on a 2d-grid and the curve represents a cut through the origin in x-direction.}
\label{Psi_Approximation_Compare}
\end{figure}
Thus, we took this approximation as a basis for further calculations of the "interaction" matrix $\mathrm{\hat{\psi}}$.

By simply calculating all matrix elements of $\mathrm{\hat{\psi}}$ by means of Equation \eqref{interact_matrix_elements}, one would neglect the finite boundaries of the system, which would cause neurons close to the edge to "bleed" into empty space. This in turn would cause the solution of \eqref{NO_equil_cond_interact_matrix_operator} to contain an over-representation of high firing rates, since close-to-the-edge neurons would need to compensate for their lack of neighbours. Thus, we had to account for the boundary conditions used in the network simulation. As described in section \ref{methods}, we simulated the network mostly with Neumann boundary conditions as well as periodic boundaries. Both types can be modelled by extending the neurons' population through spatially shifted and mirrored versions of the base population (see Figure \ref{Bound_Cond_Patches}):
\begin{itemize}
\item Periodic boundary conditions are induced by copies of the neurons' positions shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$.
\item Zero flux through the boundaries can be achieved by copied positions being shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$ \textit{and} mirrored in $x$($y$)-direction if $n_x$($n_y$) is odd.
\end{itemize}
\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{../../plots/Boundary_Cond_Sketch.png}
\end{center}
\caption{Sketch for patches of copied (and mirrored) positions for periodic (A) and Neumann (B) boundary conditions.}
\label{Bound_Cond_Patches}
\end{figure}
Therefore, the entries of the operator for periodic boundary conditions $\hat{\psi}_{per.}$ could be calculated by
\begin{equation}
\psi_{ij,per.} = \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i} + L \cdot (n_x,n_y))) \; .
\label{Psi_entries_periodic_bound}
\end{equation}
For the Neumann boundary condition, one finds
\begin{align}
\psi_{ij,neum.} &= \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} \left( d(\mathbf{x}_{neur,i},M(\mathbf{x}_{neur,i}) + L \cdot (n_x,n_y)) \right) \label{Psi_entries_neumann_bound} \\
M(\mathbf{x}) &\equiv  
 \begin{pmatrix}
  (-1)^{n_x} & 0\\
  0 & (-1)^{n_y}
 \end{pmatrix}
\mathbf{x} + L \cdot (mod(\left| n_x\right|,2),mod(\left| n_y\right|,2)) \; . \label{Mirror_Operator}
\end{align}
Note that the shift was applied to the second position. Theoretically, $n_x$,$n_y$ is iterated over all integers and in this case it is irrelevant whether to shift the first or the second position. For numerical calculations, we obviously had to limit the amount of elements of the sum such that copies that are shifted further away can be neglected.
\subsubsection{Comparison of the Solution of the Random Matrix Equation and the Simulation of the Spiking Network}

>>>>

Equation \eqref{simple_NO_dyn_with_diff_equil} can be rewritten as
\begin{equation}
\left(\nabla^2 + \left( i\sqrt{\frac{\lambda}{D}}\right)^2\right) NO = \sum_{i} \delta^2(\mathbf{x}-\mathbf{x}_{neur,i})\cdot \frac{- \gamma \cdot r_i}{D}
\label{simple_NO_dyn_with_diff_equil_helmholtz}
\end{equation}
which is a two-dimensional Helmholtz equation with a superposition of (scaled) Dirac-functions. Thus, the solution of $\mathrm{NO}$ is composed of a superposition of shifted and scaled versions of the Green's function of the differential operator on the left of the equation. For each delta function $\delta^2(\mathbf{x}-\mathbf{x}_i)$, the solution is
\begin{equation}
NO_i(\mathbf{x}) = \frac{r_i \gamma}{2 \pi D} K_0 \left(|\mathbf{x}-\mathbf{x}_{neur,i}|\sqrt{\frac{\lambda}{D}} \right) \equiv r_i \cdot \psi_{point}(|\mathbf{x}-\mathbf{x}_{neur,i}|)
\label{solution_diff_equil_bessel}
\end{equation}
where $\mathrm{K_0}$ is the zeroth modified Bessel function of the second kind \cite{Helmholtz_Solution_2d}. This solution reveals a fundamental problem of modelling the sources of $\mathrm{NO}$-production as point sources: the fact that $\mathrm{K_0(x)}$ diverges to infinity for $\mathrm{x\rightarrow 0}$. It is merely due to the finite density of the numeric grid used for the simulation of the diffusion that allows for a finite target value of concentration. Note that this problem only occurs in the two- or three-dimensional version of the differential equation, whereas in one dimension, the fundamental solution can be expressed as an exponential function with respect to the distance to the origin, resulting in a well-defined finite value at $\mathrm{x=0}$.

Generally speaking, no matter how the actual shape of the numeric solution in the equilibrium at a constant production rate looks like, it is expected to be of the form
\begin{align}
NO_i(\mathbf{x}) &= r_i \cdot \psi (d(\mathbf{x}_{neur,i},\mathbf{x})) \label{general_diff_interaction} \\
d(\mathbf{x},\mathbf{y}) &\equiv |\mathbf{x}-\mathbf{y}| \; . \label{eucl_dist}
\end{align}
The full solution is then
\begin{equation}
NO(\mathbf{x}) = \sum_i NO_i(\mathbf{x})\;.
\label{full_sol_diff_equil}
\end{equation}
If we define
\begin{equation}
\psi_{ij} \equiv \psi_{ij} \equiv \psi (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i}))
\label{interact_matrix_elements}
\end{equation}
we can express the condition \eqref{NO_equil_cond} as
\begin{equation}
\sum_j \psi_{ij}\cdot r_j = NO_0
\label{NO_equil_cond_interact_matrix}
\end{equation}
or, as an operator
\begin{align}
\hat{\psi}\mathbf{r} &= NO_0 \mathbf{n} \label{NO_equil_cond_interact_matrix_operator} \\
\mathbf{n}&\equiv (1,1,...,1) \; .
\end{align}
The problem of finding the steady-state solution of the homeostatic constraint thus reduces to inverting $\hat{\psi}$:
\begin{equation}
\mathbf{r} = NO_0 \hat{\psi}^{-1} \mathbf{n}
\label{NO_euqil_cond_interact_matrix_operator_solve}
\end{equation}
Still, to acquire any prediction from this model, we had to find a modified, non-diverging version of $\mathrm{\psi (d(\mathbf{x}_{neur,i},\mathbf{x}))}$ that retains the shape given by \eqref{solution_diff_equil_bessel} for larger distances but approaches the correct "numeric error"-value at the origin, determined by the spacing of the numeric grid. We "solved" this problem by the following expression:
\begin{equation}
\psi_{approx.} \equiv \frac{1}{\left(\frac{1}{\psi_0^\varepsilon} + \frac{1}{\psi_{point}^\varepsilon}\right)^{\frac{1}{\varepsilon}}}
\label{Numeric_Solution_Expression_Trick}
\end{equation}
where $\mathrm{\varepsilon}$ determines the "smoothness" of transition between $\mathrm{\psi}$ and the cutoff value $\mathrm{\psi_0}$. We chose $\varepsilon=10$ for all further calculations. To find an expression for $\mathrm{\psi_0}$, we took the simple approach to interpret this value as a mean of the analytic solution across the square area covered by the corresponding grid cell. As an additional simplification, we substitute the necessary integration over the square grid cell by a circular area of equal size around the source. This calculation yields
\begin{equation}
\psi_0 = \gamma \frac{1-h\sqrt{\frac{\lambda}{\pi D}} K_1\left( h\sqrt{\frac{\lambda}{\pi D}}\right) }{h^2 \lambda}
\label{Numeric_Grid_Bessel_Approx}
\end{equation}
where $\mathrm{h}$ is the spatial resolution of the grid cells. Figure \ref{Psi_Approximation_Compare} shows a comparison between the numerically calculated solution and the expression given by \ref{Numeric_Solution_Expression_Trick}. As expected, the approximation fits very well for nonzero values. The value at the critical point at the origin is slightly underestimated by \eqref{Numeric_Grid_Bessel_Approx}, but nonetheless fits well into the overall shape.
\begin{figure}
%% compare_num_diff_steady_vs_analytic.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/Psi_Interact_Approximation.png}
\end{center}
\caption{Numerical steady-state solution of Equation \eqref{simple_NO_dyn_with_diff_equil_helmholtz} (blue, $\mathrm{h=10\;\mu m}$) and its approximation (dashed) given by \eqref{Numeric_Solution_Expression_Trick}. The numerical simulation was carried out on a 2d-grid and the curve represents a cut through the origin in x-direction.}
\label{Psi_Approximation_Compare}
\end{figure}
Thus, we took this approximation as a basis for further calculations of the "interaction" matrix $\mathrm{\hat{\psi}}$.

By simply calculating all matrix elements of $\mathrm{\hat{\psi}}$ by means of Equation \eqref{interact_matrix_elements}, one would neglect the finite boundaries of the system, which would cause neurons close to the edge to "bleed" into empty space. This in turn would cause the solution of \eqref{NO_equil_cond_interact_matrix_operator} to contain an over-representation of high firing rates, since close-to-the-edge neurons would need to compensate for their lack of neighbours. Thus, we had to account for the boundary conditions used in the network simulation. As described in section \ref{methods}, we simulated the network mostly with Neumann boundary conditions as well as periodic boundaries. Both types can be modelled by extending the neurons' population through spatially shifted and mirrored versions of the base population (see Figure \ref{Bound_Cond_Patches}):
\begin{itemize}
\item Periodic boundary conditions are induced by copies of the neurons' positions shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$.
\item Zero flux through the boundaries can be achieved by copied positions being shifted by $L\cdot (n_x,n_y)$, $n_x,n_y \in \mathbb{Z}$ \textit{and} mirrored in $x$($y$)-direction if $n_x$($n_y$) is odd.
\end{itemize}
\includegraphics[width=0.5\textwidth]{../../plots/Boundary_Cond_Sketch.png}
\caption{Sketch for patches of copied (and mirrored) positions for periodic (A) and Neumann (B) boundary conditions.}
\label{Bound_Cond_Patches}
Therefore, the entries of the operator for periodic boundary conditions $\hat{\psi}_{per.}$ could be calculated by
\psi_{ij,per.} = \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} (d(\mathbf{x}_{neur,i},\mathbf{x}_{neur,i} + L \cdot (n_x,n_y))) \; .
\label{Psi_entries_periodic_bound}
For the Neumann boundary condition, one finds
\psi_{ij,neum.} &= \sum_{n_x,n_y \in \mathbb{Z}} \psi_{approx.} \left( d(\mathbf{x}_{neur,i},M(\mathbf{x}_{neur,i}) + L \cdot (n_x,n_y)) \right) \label{Psi_entries_neumann_bound} \\
M(\mathbf{x}) &\equiv  
 \begin{pmatrix}
  (-1)^{n_x} & 0\\
  0 & (-1)^{n_y}
 \end{pmatrix}
\mathbf{x} + L \cdot (mod(\left| n_x\right|,2),mod(\left| n_y\right|,2)) \; . \label{Mirror_Operator}
Note that the shift was applied to the second position. Theoretically, $n_x$,$n_y$ is iterated over all integers and in this case it is irrelevant whether to shift the first or the second position. For our calculations, we obviously had to limit the amount of elements of the sum such that copies that are shifted further away can be neglected.
\subsubsection{Comparison of the Solution of the Random Matrix Equation and the Simulation of the Spiking Network}
After having worked out the analytical basis, we compared the prediction obtained from numerically solving \eqref{NO_equil_cond_interact_matrix_operator} for certain spatial configurations of neurons to the steady-state firing rates of the full spiking network with the same spatial structure. In particular, we were interested in the quality of the predictions in respect to the choice of diffusion constant. Figure \ref{Rand_Matrix_Sol_vs_Sim} shows two examples: For $\mathrm{D=10\, \mu m^2 /s}$, the correlation between measured and predicted firing rates is very good. In contrast, we included the - obviously unsuccessful - attempt to predict firing rates for a simulation with instant diffusion based on the spatial structure by setting D to a relatively high value of $\mathrm{D=100\, \mu m^2 /s}$. Since instant diffusion overrides any spatial inhomogeneities, this represents one limiting case where our analytic model cannot make any predictions. The other limit shown in the plot, $\mathrm{D=0}$, is correctly predicted by the analytic model. However, because of the singularity of the solution, we could not quantify the correlation. Still, both extremes share their property of overriding the effect of spatial heterogeneity onto the firing rates.

\begin{figure}
%% plot_rand_matrix_sol_vs_sim.py
\includegraphics[width=\textwidth]{../../plots/rates_matrix_prediction_vs_sim.png}
\caption{Measured firing rates (t=1200-1500s) versus predicted firing rates based on the solution of Equation \eqref{NO_equil_cond_interact_matrix_operator}. For $\mathrm{D=\infty}$ (instant diffusion in the full simulation), the analytic prediction was calculated with a comparably "large" diffusion constant of $\mathrm{D=100\, \mu m^2 /s}$.}
\label{Rand_Matrix_Sol_vs_Sim}
\end{figure}
